---
title: "Tutorial - mvHMM"
author: "Beniamino Hadj-Amar"
date: '2023-09-20'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Working directory

+ Change this to be the directory that '/include' is saved in 

```{r setwd, include=TRUE,echo=TRUE, eval=TRUE,cache=TRUE}

my_dir <- "~/Desktop_New/Research/Neuroscience/pupillometry_new/published_code/"
setwd(my_dir)
```


## Preamble and Settings {.tabset}

### Packages

Loading the required packages.

```{r packages, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE, message = FALSE, results = FALSE}

library("R.matlab")
library("rstan")
library("plyr")
library("dplyr")
library("magrittr")
library("bayesplot")
library("bridgesampling")
library("matrixStats")
library("mvtnorm")
library("reshape2")
library("timeSeries")
library("stringr")
library("RColorBrewer")
library("tibble")
library("pdftools")
library("fabricatr")
library("lomb")
library("writexl")
library("imputeTS")


stan_path <- "stan/HMM_MvNormEmis.stan" # stan path mvHMM
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

```

### Utility Functions

Calling utility functions
```{r utils, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}

source("include/util_data.R")
source("include/util_mvHMM.R")
source("include/util_stan.R")

```

### Auxiliary Settings

Auxiliary settings for plots

```{r auxiliary_plot_stuff, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
# - plt vars
plt_pars <- list()
plt_pars$pch <- c(18, 20)
plt_pars_joint <- list()
plt_pars_joint$pch <- c(20, 20, 20)
plt_pars_joint$xlab <- rep("Time (Seconds)", 2)
plt_pars_joint$ylab <- c("NE", "Pupil")
plt_pars_joint$zcol <- brewer.pal(n = 4, name = "Spectral")
neuromods  = c("NE", "Pupil")
plt_pars_joint$neuromods <- neuromods
corr_label <- paste("corr(", str_c(neuromods, 
                                   collapse = ","), ")", sep = "")
# info variables
info_vars <- list()
info_vars$model <- "Epilepsy Monitoring Unit"
info_vars$id <- c(123)
info_vars$id_aux <- c("CC-1,2,3")
info_vars$disease <- rep("EMU", length(info_vars$id))
info_vars$nm <- c("NE", "Pupil") 
info_vars$brain_region <- rep("Amygdala", length(info_vars$id))
# - 
```

### Select subject, available snippets, and substrats

Select subject and Available snippets and substratifications
```{r select_substrat, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
ID = 123 # average of sbj 1,2, and 3. 
# snippets <- c("stimulus")
snippet = "stimulus"
substrat = "Oddball_x_Arousal"
substrat_groups <- c("Oddball_ArousLow","Oddball_ArousHigh")
```

### Pre-process options
Selecting window smoothing and de-trend options
```{r pre_process_option, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
win_s <- 5 # window smoothing
detrend <- TRUE # detrend time series
```

### MCMC options and hyperparms 
Selecting number of MCMC iterations and hyperparameters
```{r MCMC_option, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
n_MCMC = 5000
mu_scale = 1.5
tau_scale = 1.5 
```



## (1) Oddball and High Arousal {.tabset}

### Select substrat group
Selecting substratification (jj = 2: oddball x high arousal)
```{r OddballHighArousal_group, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
jj = 2
group = substrat_groups[jj]; cat(group)
```

### Get data + plot
```{r OddballHighArousal_plot, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
obs_group <- get_neuromodulators(ID, info_vars, snippet = "stimulus", 
                           substrat = substrat,
                           groups = substrat_groups, 
                           win_s = win_s,
                           detrend = detrend, plt = T)[[jj]]
```

### MCMC Run
Running MCMC sampler using stan model
```{r OddballHighArousal_MCMC, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
K = 3 # number of states (selected among compiting models using bridgesampling)
fit <- stan(file = stan_path,
                      data = list(N = nrow(obs_group), D = 2, 
                                  K = K, y = obs_group, 
                                  mu_loc = 0,
                                  mu_scale = mu_scale, 
                                  alpha_0 = matrix(c(rep(1, K*K)), 
                                                   nrow = K, ncol = K, byrow = TRUE),
                                  tau_loc = 1, tau_scale = tau_scale,
                                  Omega_shape = 1),  
                      seed = 123, 
                      chains = 1, iter = n_MCMC, cores = 1)
print(fit, pars = c("mu", "tau", "gamma"), 
                probs = c(.05, .95))
```

### Extract Fit + Diagnostics + Performance
Extracting fit + convergence diagnostics + performances measures
```{r OddballHighArousal_ExtractFit, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
sims <- rstan::extract(fit)
perf <- mvHMM_performance(fit, obs_group) 
```

### Posterior predictive + plot
Posterior predictive plot
```{r OddballHighArousal_postpred, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
mvHMM_predictive <- mvHMM_get_predictive(fit, obs_group, ndraw = 200)
z_hat <- mvHMM_predictive$z_hat
y_hat <- mvHMM_predictive$y_hat
mvHMM_plot_predictive_joint(obs_group, y_hat,
                                      z_hat, plt_pars_joint,
                                      snippet = snippet, win_s = win_s)
dev.off()
```

### Posterior correlation + plot
Time-varying posterior correlation plot
```{r OddballHighArousal_correlation, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
corr_sims <- mvHMM_get_correlation(sims, obs_group)
mvHMM_plot_correlation(corr_sims, z_hat, plt_pars_joint$zcol,
                       corr_label, 
                       cols_grey, snippet = snippet, substrat = "allTrials",
                       plt_state_probs = T, win_s = win_s)
dev.off()
```

### State probabilities plot
Time-varying state probability plot
```{r OddballHighArousal_stateprob, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
mvHMM_plot_stateprobs(sims, plt_pars_joint$zcol,
                      obs_group, snippet = snippet, win_s = win_s)
```


## (2) Oddball and Low Arousal{.tabset}

### Select substrat group
Selecting substratification (jj = 1: oddball x low arousal)
```{r OddballLowArousal_group, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
jj = 1
group = substrat_groups[jj]; cat(group)
```

### Get data + plot
```{r OddballLowArousal_plot, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
obs_group <- get_neuromodulators(ID, info_vars, snippet = "stimulus", 
                           substrat = substrat,
                           groups = substrat_groups, 
                           win_s = win_s,
                           detrend = detrend, plt = T)[[jj]]
```

### MCMC Run
Running MCMC sampler using stan model

```{r OddballLowArousal_MCMC, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
K = 4 # number of states (selected among compiting models using bridgesampling)
fit <- stan(file = stan_path,
                      data = list(N = nrow(obs_group), D = 2, 
                                  K = K, y = obs_group, 
                                  mu_loc = 0,
                                  mu_scale = mu_scale, 
                                  alpha_0 = matrix(c(rep(1, K*K)), 
                                                   nrow = K, ncol = K, byrow = TRUE),
                                  tau_loc = 1, tau_scale = tau_scale,
                                  Omega_shape = 1),  
                      seed = 123, 
                      chains = 1, iter = n_MCMC, cores = 1)
print(fit, pars = c("mu", "tau", "gamma"), 
                probs = c(.05, .95))
```

### Extract Fit + Diagnostics + Performance
Extracting fit + convergence diagnostics + performances measures
```{r OddballLowArousal_ExtractFit, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
sims <- rstan::extract(fit)
perf <- mvHMM_performance(fit, obs_group) 
```

### Posterior predictive + plot
Posterior predictive plot
```{r OddballLowArousal_postpred, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
mvHMM_predictive <- mvHMM_get_predictive(fit, obs_group, ndraw = 200)
z_hat <- mvHMM_predictive$z_hat
y_hat <- mvHMM_predictive$y_hat
mvHMM_plot_predictive_joint(obs_group, y_hat,
                                      z_hat, plt_pars_joint,
                                      snippet = snippet, win_s = win_s)
dev.off()
```

### Posterior correlation + plot
Time-varying posterior correlation plot
```{r OddballLowArousal_correlation, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
corr_sims <- mvHMM_get_correlation(sims, obs_group)
mvHMM_plot_correlation(corr_sims, z_hat, plt_pars_joint$zcol,
                       corr_label, 
                       cols_grey, snippet = snippet, substrat = "allTrials",
                       plt_state_probs = T, win_s = win_s)
dev.off()
```

### State probabilities plot
Time-varying state probability plot
```{r OddballLowArousal_stateprob, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
mvHMM_plot_stateprobs(sims, plt_pars_joint$zcol,
                      obs_group, snippet = snippet, win_s = win_s)
```



## (3) All Trials {.tabset}


### Get data + plot
```{r AllTrials_plot, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
obs <- get_neuromodulators(ID, info_vars, snippet = "stimulus", 
                                      substrat = "allTrials",
                                      groups = NULL, win_s = win_s,
                                      detrend = detrend, plt = T) 
```

### MCMC Run
Running MCMC sampler using stan model

```{r AllTrials_MCMC, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
K = 3 # number of states (selected among compiting models using bridgesampling)
fit <- stan(file = stan_path,
            data = list(N = nrow(obs), D = 2, 
                        K = K, y = obs, 
                        mu_loc = 0,
                        mu_scale = mu_scale, 
                        alpha_0 = matrix(c(rep(1, K*K)), 
                                         nrow = K, ncol = K, byrow = TRUE),
                        tau_loc = 1, tau_scale =tau_scale,
                        Omega_shape = 1),  
            seed = 123, 
            chains = 1, iter = n_MCMC, cores = 1)
print(fit, pars = c("mu", "tau", "gamma"), 
                probs = c(.05, .95))
```

### Extract Fit + Diagnostics + Performance
Extracting fit + convergence diagnostics + performances measures
```{r AllTrials_ExtractFit, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
sims <- rstan::extract(fit)
perf <- mvHMM_performance(fit, obs_group) 
```

### Posterior predictive + plot
Posterior predictive plot
```{r AllTrials_postpred, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
mvHMM_predictive <- mvHMM_get_predictive(fit, obs, ndraw = 200)
z_hat <- mvHMM_predictive$z_hat
y_hat <- mvHMM_predictive$y_hat
mvHMM_plot_predictive_joint(obs, y_hat,
                                      z_hat, plt_pars_joint,
                                      snippet = snippet, win_s = win_s)
dev.off()
```

### Posterior correlation + plot
Time-varying posterior correlation plot
```{r AllTrials_correlation, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
corr_sims <- mvHMM_get_correlation(sims, obs)
mvHMM_plot_correlation(corr_sims, z_hat, plt_pars_joint$zcol,
                       corr_label, 
                       cols_grey, snippet = snippet, substrat = "allTrials",
                       plt_state_probs = T, win_s = win_s)
dev.off()
```

### State probabilities plot
Time-varying state probability plot
```{r AllTrials_stateprob, include=TRUE, echo=TRUE, eval=TRUE, cache=FALSE}
mvHMM_plot_stateprobs(sims, plt_pars_joint$zcol,
                      obs, snippet = snippet, win_s = win_s)
```


